---
title: "Analysis of ToothGrowth Data"
author: "Shane Reed"
date: "November 20, 2015"
output: word_document
---

<style type="text/css">
.table {
    width: 60%;
}
</style>

##Overview
In this report I will exam the `ToothGrowth` data set from the `datasets` R library. According to the R Documentation, the `ToothGrowth` data set contains the results from an experiment in which "the response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid)." After a brief exploratory data analysis, I perform multiple hypothesis tests to see if the two factors, dose and delivery method, have a significant effect on tooth growth.

##Exploratory Data Analysis
```{r echo=TRUE, message=FALSE, warning=FALSE, fig.width=5, fig.height = 4}
library(ggplot2)
ggplot(aes(supp, len), data = ToothGrowth) + 
  geom_boxplot(aes(fill = supp)) + facet_grid(. ~ dose) + 
  labs(x = "Supplement", y = "Length", title = "Tooth Growth - OJ vs. VC by Dose")
```

After an initial exploration of the data using the box plot shown above, the following points appear to be true:

* For doses of .5 mg & 1.0 mg OJ promotes greater growth
* For the dose of 2.0 mg VC promotes greater growth
* For a given supplement, each incremental increase in dose promotes greater growth

##Hypothesis Tests and Results
###Hypothesis Tests
I will use the `t.test` function, with a 95% confidence level, to test whether the observations outlined above are true. In other words, I will use hypothesis tests, in which the null hypothesis is stated as $H_0: \mu_1 = \mu_2$, to see if there is statistically significant difference between each corresponding groups.
```{r message=FALSE, warning=FALSE}
library(dplyr)
# Compare delivery method @ .5 mg
d05 <- filter(ToothGrowth, dose == .5)
t05 <- t.test(len ~ supp, data = d05)
# Compare delivery method @ 1.0 mg
d10 <- filter(ToothGrowth, dose == 1)
t10 <- t.test(len ~ supp, data = d10)
# Compare delivery method @ 2.0 mg
d20 <- filter(ToothGrowth, dose == 2)
d20oj <- d20$len[d20$supp == "OJ"]; d20vc <- d20$len[d20$supp == "VC"]
t20 <- t.test(d20vc, d20oj)
# Compare doses for OJ: .5 mg to 1.0 mg, .5 mg to 2.0 mg, 1.0 mg to 2.0 mg
oj <- filter(ToothGrowth, supp == "OJ")
oj05 <- filter(oj, dose == .5); oj10 <- filter(oj, dose == 1); oj20 <- filter(oj, dose == 2)
tOj10_05 <- t.test(oj10$len, oj05$len)
tOj20_05 <- t.test(oj20$len, oj05$len)
tOj20_10 <- t.test(oj20$len, oj10$len)
# Compare doses for VC: .5 mg to 1.0 mg, .5 mg to 2.0 mg, 1.0 mg to 2.0 mg
vc <- filter(ToothGrowth, supp == "VC")
vc05 <- filter(vc, dose == .5); vc10 <- filter(vc, dose == 1); vc20 <- filter(vc, dose == 2)
tVc10_05 <- t.test(vc10$len, vc05$len)
tVc20_05 <- t.test(vc20$len, vc05$len)
tVc20_10 <- t.test(vc20$len, vc10$len)
```

###Delivery Method Results
The results from the first 3 tests are shown in the table below. Based on these results, we can reject $H_0$ for doses 0.5 mg and 1.0 mg, i.e. the true difference in means is not equal to 0; in both cases, the OJ delivery method promotes greater tooth growth than VC. However, we cannot reject $H_0$ for the 2.0 mg dose.

Dose|p-value|Confidence Interval|Mean (OJ)|Mean (VC)
---|---|---|---|---
0.5 mg|`r round(t05$p.value, 3)`|[`r round(t05$conf.int[1], 2)`, `r round(t05$conf.int[2], 2)`]|`r t05$estimate[[1]]`|`r t05$estimate[[2]]`
1.0 mg|`r round(t10$p.value, 3)`|[`r round(t10$conf.int[1], 2)`, `r round(t10$conf.int[2], 2)`]|`r t10$estimate[[1]]`|`r t10$estimate[[2]]`
2.0 mg|`r round(t20$p.value, 3)`|[`r round(t20$conf.int[1], 2)`, `r round(t20$conf.int[2], 2)`]|`r t20$estimate[[1]]`|`r t20$estimate[[2]]`

###Dose Level Results
The results from the remaining 6 tests are shown in the table below. Based on these results, we can reject $H_0$ for all of the test. 

Delivery|X|Y|p-value|Confidence Interval|Mean (X)|Mean (y)
---|---|---|---|---|---|---
OJ|1.0 mg|0.5 mg|`r format(signif(tOj10_05$p.value, 3), scientific = T)`|[`r round(tOj10_05$conf.int[1], 2)`, `r round(tOj10_05$conf.int[2], 2)`]|`r tOj10_05$estimate[[1]]`|`r tOj10_05$estimate[[2]]`
OJ|2.0 mg|0.5 mg|`r format(signif(tOj20_05$p.value, 3), scientific = T)`|[`r round(tOj20_05$conf.int[1], 2)`, `r round(tOj20_05$conf.int[2], 2)`]|`r tOj20_05$estimate[[1]]`|`r tOj20_05$estimate[[2]]`
OJ|2.0 mg|1.0 mg|`r format(signif(tOj20_10$p.value, 3), scientific = T)`|[`r round(tOj20_10$conf.int[1], 2)`, `r round(tOj20_10$conf.int[2], 2)`]|`r tOj20_10$estimate[[1]]`|`r tOj20_10$estimate[[2]]`
VC|1.0 mg|0.5 mg|`r format(signif(tVc10_05$p.value, 3), scientific = T)`|[`r round(tVc10_05$conf.int[1], 2)`, `r round(tVc10_05$conf.int[2], 2)`]|`r tVc10_05$estimate[[1]]`|`r tVc10_05$estimate[[2]]`
VC|2.0 mg|0.5 mg|`r format(signif(tVc20_05$p.value, 3), scientific = T)`|[`r round(tVc20_05$conf.int[1], 2)`, `r round(tVc20_05$conf.int[2], 2)`]|`r tVc20_05$estimate[[1]]`|`r tVc20_05$estimate[[2]]`
VC|2.0 mg|1.0 mg|`r format(signif(tVc20_10$p.value, 3), scientific = T)`|[`r round(tVc20_10$conf.int[1], 2)`, `r round(tVc20_10$conf.int[2], 2)`]|`r tVc20_10$estimate[[1]]`|`r tVc20_10$estimate[[2]]`

##Conclusions and Assumptions
At the .05 mg and 1.0 mg doses, OJ is promotes greater tooth growth than VC; however, at a 2.0 mg dose we cannot say that one delivery method results in more growth than the other.  Furthermore, when we look at incremental doses, using the same delivery method, higher doses promote more growth than lower doses.  These conclusions, and the test performed to arrive at these conclusions, are based on the following assumptions:

* The guinea pigs were chosen using an accepted method for randomization
* The sample groups are independent
* The variances of each group, in any given test, are not equal
